---
title: "PS270 Final Project"
author: Julia Wright
description: "My Final Project"
output:
  distill::distill_article:
    self_contained: false
---

## Project thoughts

I am interested in exploring data related to...

I am interested looking at election outcomes/polling data to see if predictions based on polling data before elections translates to the outcome. I would also be interested in looking into campaign finance similar to problem set 2. I thought it was interesting looking at which candidates spent the most on advertising. 

## Milestone 2, data and tentative research question
I will be using the dataset “U.S. President Precinct-Level Returns 2020” from Harvard Dataverse to examine if certain modes of voting, such as absentee and mail in or in-person on Election Day, correlates to the candidate/party that received the votes. 

My question will be does the mode of voting have an effect on the votes given to the Democratic Party and its candidates? 

I hypothesize that a greater amount of absentee votes (mail in votes) are received by the Democratic Party. I hypothesize this due to behavioral patterns that I observed during the 2020 elections and the COVID-19 pandemic. I observed that many Democrats voted via mail-in ballot to avoid the potential of spreading COVID-19. Republicans seemed more comfortable going to the polls in person, and debated the legitimacy of mail-in ballots. These behavioral observations lead me to hypothesize that more absentee votes are Democrat votes. 

The explanatory variable of interest will be the form or more of voting. This is measured by the category present in the data which is either absentee or normal/in-person. This has been pre-recorded. 

The outcome variable will be which party was voted for. It will either be a democratic or republican candidate, with some writing in and voting for a third party. I will be looking at the share of votes received by the democrats that are mail-ins/absentee. 

If a large majority of absentee votes were received by the Democrats, this could provide support for my hypothesis. If it does not seem that most of the mail-in votes were received by Democrats, meaning more Republicans or third-party voters were mail-ins, then my hypothesis will be false. 

## Data Visualization

```{r}
data <- read.csv("PRESIDENT_precinct_reduced.csv")

library(tidyverse)

data <- data[, !(names(data) %in% c("state", "party_detailed", "votes", "year", "precinct", "office", "county_name", "county_fips", "jurisdiction_name", "jurisdiction_fips", "candidate", "district", "dataverse", "stage", "special", "writein", "state_po", "state_fips", "state_cen", "state_ic", "date", "readme_check", "magnitude"))]

write.csv(data, "PRESIDENT_precinct_reduced.csv", row.names = FALSE)

file.info("PRESIDENT_precinct_reduced.csv")$size 


```

## Graph 
```{r fig.width= 20, fig.height = 4}
library(tidyverse)


filter_data <- data |> filter((party_simplified == "DEMOCRAT" | party_simplified == "REPUBLICAN" | party_simplified == "OTHER") &
         (mode == "ABSENTEE" | mode == "NOT ABSENTEE"))

ggplot(filter_data, aes(x = mode, fill = party_simplified)) +
  geom_bar(position = "dodge") +
  labs(title = "Voting mode by party",
       x = "Voting Mode (Absentee vs. Not Absentee or In Person)",
       y = "Number of votes by party",
       fill = "Party") +
  theme_minimal()
print(ggplot)
```

## Second Visualization and Analysis

```{r}
library(tidyverse)
library(broom)

data <- data |>
  filter(party_simplified  == "DEMOCRAT"
         | party_simplified == "REPUBLICAN"
         | party_simplified == "OTHER") |>
  mutate(dem_voters = ifelse(party_simplified == "DEMOCRAT", 1, 0),
        absentee_votes = ifelse(mode == "ABSENTEE", 1, 0))

table <- lm(dem_voters ~ absentee_votes, data = data)

table |>
  broom::tidy() |>
  select(term, estimate) |>  
  knitr::kable(digits = 2)
```
```{r}
ggplot(data, aes(x = absentee_votes, y = dem_voters)) +
  geom_point(color = "steelblue")
```

The intercept of 0.17 means that when absentee votes is 0, about 17% of votes that are in person and at the polls will be for the Democratic candidate. 

Since the coefficient for absentee votes is 0.00, this means that the mode of voting should have no effect on whether a voter votes for a Democratic or Republican candidate. 

These findings contradict and disprove my hypothesis that more Democratic voters would be voting via mail in ballots versus in person, directly at the polls. 

